Pool Model Paper

### 2023-08-10 pool_model.tex
add text about cell death

@@

Consider for simplicity the system where only one bacteria species is present $N=1$.
According to the growth curves (\cmt{some references on bacteria growth curves??}) quite often at the beginning no increase in bacterial count is present and so-called lag-phase is observed.
This phenomenon appears due to adaptation of the microbial community, for example, to the new environmental conditions.
After this some bacteria enter the exponential growth phase, while some of it may not be able to adjust and experience the cell death.
Some bacteria even may not be able to adjust and die other will enter the exponential growth phase after a lag phase. 
After exponential growth the system reaches the saturation where the balance between the number of dying bacteria is in balance with the growing one leads to the stationary state.
In the described simplified version of the real-life system three possible bacteria states are mentioned corresponding to which process is happeng: adaplation, growth or death.
Motivated by this we divide a bacterial population into three pools (states) (\textbf{Figure \ref{SchematicRep}}): 
\begin{itemize}
\item $L(t)$: the fraction of the population in the lag phase
\item $G(t)$: the fraction of the population growing and dividing
\item $D(t)$: the fraction of the population undergoing cell death
\end{itemize}
Here the concentrations of the pools are components of our variable vector $\mbx (t) =\{L(t), G(t), D(t)\}$.

\begin{figure}[t]
    \begin{center}
    \includegraphics[width=0.7\textwidth]{Figures/TPM_fig.pdf}
    \caption{Schematic representation of exchange between three states or pools.}
    \label{SchematicRep}
    \end{center}
    \end{figure}

@@

Based on experimental observations, it was shown that the bacterial growth is strongly influenced by the limiting resource. \cmt{ref review}
In the system where all needed nutritients are in excess except for one the total growth is linear to the conentration of this liminting nutritient. \cmt{ref Monod}
\cmt{Ref about Quorum sensing. Signalling mechanism that controll gene expression (bacteria growth within species) in response to environment. }

According to this, we aim to include a resource pool which is used by the growing bacteria using the second order kinetic reaction:
\begin{equation}
    G + R_0  \stackrel{\alpha_0}{\longrightarrow} 2G.
\end{equation}

@@

@@

As it was mentioned above, bacteria usually coexist in large multispecies interacting communities whose growth is vastly determined by the developed survival mechanisms.
These survival mechanisms mostly aim to secure as much available nutrition which lead to competitive strategies or, if it is advantageous, cooperative ones.
Further we show that the TPM can be readily extended to the systems with more than one bacteria an take describe more complex relationships between species.

@@

One on the ways to descibe competition between bacteria populations is as a function of limiting resources ratio. \cmt{Ref Tilman 1977, Ref Smith 2002}
To descibe such a situation we consider for simplicity only one common limiting resource pool. 
The simplest pool model for bacterial species $A$ and $B$ encompassing lag-phases, growth-phases and a common resource pool is given by (no back-flow to the lag-pool):

@@

which has the structure of a generalised Lotka-Volterra system. In order to better see this, we set $L_A=L_B=0$: 
\begin{eqnarray}
\label{Comp_Model}
\dot{G}_A &=& \alpha_A G_A\left(1 - \frac{G_A}{N_t}\right ) - \frac{\alpha_A}{N_t}G_AG_B\\
\dot{G}_B &=& \alpha_B G_B\left(1-\frac{G_B}{N_t}\right ) -\frac{\alpha_B}{N_t}G_AG_B. 
\end{eqnarray}

@@

@@

@@

On the other hand, one can also observe some cooperative behaviours between species in case when it facilitates the competition with others.
Quite often it happens between the species that are close relatives or clonal populations and such process is called kin selection. \cmt{Ref 25 from review}
Sometimes neighbours, so-called 'social cheaters', can benefit from a certain behaviour without expressing it by itself.  \cmt{Ref 29 from review}

@@

@@

>>>>

\cmtb{Maybe add here motivations about bacteria states: So toy system where we can eliminate states (sum and get F(n1, n2)).
And describe it initially without states/1state. Get the difference.}

@@

Consider for simplicity the system where only one bacteria species is present $N=1$.
The bacterial growth curves usually consists of different consecutive phases~\cite{buchanan_when_1997}.
At the beginning no increase in bacterial count is present and so-called lag-phase is observed.
This phenomenon appears due to adaptation of the microbial community, for example, to the new environmental conditions.
This so-called preparation to the exponential growth includes synthesys of the needed compounds, initiation of the transciption~\cite{rolfe_lag_2012} and poorly studied physiological and regulatory mechanisms~\cite{monod_growth_1949,}
\cmt{For example, in Baranyi and Roberts modelling of lag-phase ???(maybe add).}
After this, the succeeded in adaptation bacteria enter the exponential growth phase, where the process of cell division is directly happening \cmt{Add Ref}.
When the nutrition is exhausted by the growing bacteria, the stationary state of the growth curve occurres.
In this state the cultures accumulate waste products of metabolitic activity that leads to cell death of the bacteria.
At the same time dying cells release the nutritions for survivors to continue growing.
As a result the balance between the growing and dyuing processes leads to the constant with time bacterial count~\cite{navarro_llorens_stationary_2010}.
In the described simplified version of the real-life system three possible bacteria states are mentioned corresponding to which process is happeng: adaplation, growth or death.
Motivated by this we divide a bacterial population into three pools (states) (\textbf{Figure \ref{fig:SchematicRep}}): 
\begin{itemize}
\item $L(t)$: the fraction of the population in the lag phase
\item $G(t)$: the fraction of the population growing and dividing
\item $D(t)$: the fraction of the population undergoing cell death
\end{itemize}
Here the concentrations of the pools are components of our variable vector $\mbx (t) =\{L(t), G(t), D(t)\}$.

\begin{figure}[t]
    \begin{center}
    \includegraphics[width=0.7\textwidth]{Figures/TPM_fig.pdf}
    \caption{Schematic representation of exchange between three states or pools.}
    \label{fig:SchematicRep}
    \end{center}
    \end{figure}

@@

Based on experimental observations, it was shown that the bacterial growth is strongly influenced by the limiting resource.
In the system where all needed nutritients are in excess except for one the total growth is linear to the conentration of this liminting nutritient~\cite{hibbing_bacterial_2010, monod_growth_1949}.
The controll of the population density with respect to the available nutritions is implemented using cell communication process called quorum sensing. 
Using release of the extracellular signalling molecules (autoinducers) it is possible to synchronize the gene expression of the whole group~\cite{ng_bacterial_2009}.

We would not dive into the processes of quorum sensing and try to keep it as simple as possible.
According to this, we aim to include a resource pool which is used by the growing bacteria using the second order kinetic reaction:
\begin{equation}
    G + R_0  \stackrel{\alpha_0}{\longrightarrow} 2G.
\end{equation}

@@

\cmt{Look into this!}

@@

As it was mentioned above, bacteria usually coexist in large multispecies interacting communities whose growth is vastly determined by the developed survival mechanisms.
These survival mechanisms mostly aim to secure as much available nutrition which lead to competitive strategies or, if it is advantageous, cooperative ones~\cite{hibbing_bacterial_2010, stubbendieck_bacterial_2016}.
Further we show that the TPM can be readily extended to the systems with more than one bacteria an take describe more complex relationships between species.

@@

One on the ways to descibe competition between bacteria populations is as a function of limiting resources ratio~\cite{tilman_resource_1977, smith_effects_2002}.
To descibe such a situation we consider for simplicity only one common limiting resource pool. 
The simplest pool model for bacterial species $A$ and $B$ encompassing lag-phases, growth-phases and a common resource pool is given by (no back-flow to the lag-pool):

@@

which has the structure of a generalised Lotka-Volterra system. \cmt{Add reference about Lotka-Volterra. Compare with different states LV.}
In order to better see this, we set $L_A=L_B=0$: 
\begin{eqnarray}
\label{Comp_Model}
\dot{G}_A &=& \alpha_A G_A\left(1 - \frac{G_A}{N_t}\right ) - \frac{\alpha_A}{N_t}G_AG_B\\
\dot{G}_B &=& \alpha_B G_B\left(1-\frac{G_B}{N_t}\right ) -\frac{\alpha_B}{N_t}G_AG_B. 
\label{eq:LV_simple}
\end{eqnarray}

Equations (\ref{eq:LV_simple}) are basically competitive Lotka-Volterra equations that are widely used in ecological studies and show the principle of competive excursion.
That principle claims that two species in competition for the same resource cannot coexist~\cite{wangersky_lotka-volterra_1978}.

@@

\cmtb{Look papers on modelling of the LV in bacterial communities. Did anyone do the same as we??}

@@

\cmt{\cite{lenski_coexistence_1986}: Modelling of the 2 sp. system with 1resource and 1 inhibitor.}

@@

On the other hand, one can also observe some cooperative behaviours between species in case when it facilitates the competition with others.
Quite often it happens between the species that are close relatives or clonal populations and such process is called kin selection \cite{west_social_2007}.
Sometimes neighbours, so-called 'social cheaters', can benefit from a certain behaviour without expressing it by itself \cite{rainey_evolution_2003}.

\cmt{\cite{rocha_base_2002}: Symbiosis host/bacteria. bacterial genomes suffering reductive evolution due to this. 
Maybe it's connected with 'social cheaters'?}

@@

\section{Spatial Limitations}

\cmtb{Maybe as motivation can be spatial competitive mechanisms as adhesion to survaces and motility. 
So how much including these effects will affect system.}

\cmt{\cite{htuson_bacteriasurface_2013}: Attachment of bacteria to surfaces. (One of the mechanisms to stick to the areas with nutritients)}

\cmtb{We can take some toy model (e.g the one from group seminar) and do agent-based simulation. (add spatial component)
Then from this simulation we sum bacteria and get growth curves.
And using our model equations we get also growth curves.
And find prob diffusion coef, maybe rates where model works fine.
For sim maybe we will need to name specific bacteria species. (maybe nicer)}

@@

\newpage
\printbibliography
